---
title: "Assignment 5 - Factors and data input-output"
author: "Trevor Baker"
date: "15/10/2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, message = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(gapminder)
library(forcats)
library(here)
library(DT)

```

<!--This is to set up the two-column and clearer output formats-->
<style type="text/css">
.twoC {width: 100%}
.clearer {clear: both}
.twoC .table {max-width: 50%; float: right}
.twoC img {max-width: 50%; float: left}
</style>
<!-- end two-column setup -->


## Exercise 1: The value of the here::here package

Setting working directories is not as straightforward as it seems. If someone else wants to run code that has its working directory defined with `setwd()`, then they will need to have the same folder structure as the originating computer. Likewise if you get a new computer, or move your work to another folder. There are additional compatibility problems when switching between operating systems as some use `\\` to separate folders in file paths, and others use `/`. There are additional inconsistenciegs between how `.R` files and `.Rmd` files deal with working directories.

It is a big hassle to have to overwrite working directories manually within scripts, especially if `setwd()` is called numerous times.

The `here` package provides a solution to these issues by providing a platform-agnostic way to set working directories and define file paths. With `here`, working directories and file paths are set relative to each user's file structure and operating system. The working directory when using `here` is defined based on simple rules, such as where a `.here` file was made using the function `here::set_here()`, or where an `.Rproj` file exists in the current folder structure. By opening projects via `.Rproj` files and using `here::here` within scripts, any computer access the script via the `.Rproj` file will be able to seamlessly access and write files within that script using the folder structure and operating system that exists locally. 

_`here` makes work easily transferable, and readily reproducible, which should be a goal in all data analyses._


## Exercise 2: Factor management

#### Gapminder option

__Drop Oceania.__ _Filter the Gapminder data to remove observations associated with the continent of Oceania. Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and Oceania; address the number of rows and the levels of the affected factors._

The code chunk below ensures that the `gapminder$continent` column is indeed a factor:
```{r factor_drop_check}

#ensure that the column is a factor
gapminder$continent %>% class()

```

The code chunk below provides a look at factor levels in `gapminder$continent`before Oceania is dropped. There are actually two factor columns in `gapminder`: continent and country. If we were dropping Oceania from the dataset for further analysis, it would be important to also remove Oceania-associated factor levels from the country column, as they will have no associated obervations (i.e. rows) remaining, so these will be dealt with as well in the following steps. 
```{r factor_drop_before}

#look at unique levels of gapminder$continent before removing Oceania:
gapminder$continent %>% levels()

#we have 5 levels in the continent column beforehand:
gapminder$continent %>% nlevels()

#there are also Oceania-associated factor levels (i.e. countries) in gapminder$country:
gapminder %>%
  group_by(continent) %>%
  summarize(n_country = n_distinct(country))
  
#these two countries in Oceania are Australia and New Zealand:
gapminder %>%
  filter(continent=="Oceania") %>%
  select(country) %>%
  unique()

```


This chunk of code removes the Oceania rows from the dataset using the `filter()` function, and then drops all unused levels from all factor columns (i.e. country and continent) using the `droplevels()` function:
```{r factor_drop_filter}

#this is the step to drop Oceania and associated factor levels from the dataset
gapminder_drop <- gapminder %>%
                    filter(continent != "Oceania") %>%
                    droplevels()

```


This last code chunk from this task shows in multiple ways that Oceania has been dropped, as have associated factor levels from both the country and continent columns:
```{r factor_drop_after}

#this shows that now Oceania is not one of the levels in the continent column:
gapminder_drop$continent %>% levels()

#and this code outputting zero rows further confirms that Oceania is no longer in the dataset:
gapminder_drop %>% filter(continent == "Oceania")

#this shows that there are no more countries in the dataset from the Oceania continent:
gapminder_drop %>%
  group_by(continent) %>%
  summarize(n_country = n_distinct(country))
  
#and this code shows that the factor levels for Australia and New Zealand are no longer in the country column (i.e. the length of the number of country levels which are equal to Australia or New Zealand is zero):
gapminder_drop$country %>% 
  levels() %in% c("Australia", "New Zealand") %>%
  which() %>%
  length()

```

If, for some reason, you only wanted to drop the Oceania factor level from the continent column but keep the now-empty factor levels in the country column, you could use `forcats::fct_drop()` as in the code chunk below:
```{r factor_drop_forcats}

#This drops Oceania from the dataset and removes only unused factor levels in the continent column:
gapminder_drop2 <- gapminder %>%
                    filter(continent != "Oceania") %>%
                    mutate(continent = forcats::fct_drop(continent))

#now we can see that Oceania is no longer present in the continent factor levels:
gapminder_drop2$continent %>% levels()

#but Australia and NZ can still be found in the country column's factor levels (i.e. the length of the number of country levels which are equal to Australia or New Zealand is 2):
gapminder_drop2$country %>% 
  levels() %in% c("Australia", "New Zealand") %>%
  which() %>%
  length()

#this is despite the Australia and NZ rows no longer being part of the dataset, as shown by this code returning zero rows:
gapminder_drop2 %>% 
  filter(country %in% c("Australia", "New Zealand"))


```


__Reorder the levels of country or continent.__ _Use the forcats package to change the order of the factor levels, based on summarized information of one of the quantitative variables. Consider experimenting with a summary statistic beyond the most basic choice of the mean/median. Use the forcats package in the tidyverse for this, rather than the baseR function as.factor._

Reordering factor levels is helpful if you are making graphs with a categorical (i.e. factor-based) axis and want a sensible order for display. Without doing this, the graphed summary data would be displayed in disarray. For this task, I have chosen to reorder and display data by the magntiude of their coefficients of variation (CV) in per-capita GDP. 

In order to do this, I first had to make a simple function to compute CV, which is equal to the standard deviation divided by the mean.
```{r coeff_var}

#define coefficient of variation function, round result to make it easier to read in datatable
coeff_var <- function(x){ (sd(x) / mean(x)) %>% round(digits=4) }

```

In the code chunk below, the order of the country factor levels is changed based on the coefficent of variation in per-capita GDP observed for each country. I also made a new column holding the country column as it was given (country_asGiven), in order to use it for the example graphs below.

In order to do this, I first had to make a simple function to compute the CV, which is equal to the standard deviation divided by the mean.
```{r factor_reord}

#reorder the factor according to the coefficent of variation
gm_reord <- gapminder %>%
              mutate(country_asGiven = country) %>% #this line just duplicates the country column as given, preserving the old factor order
              mutate(country = forcats::fct_reorder(country, gdpPercap, coeff_var)) %>%#this line reorders the country factor levels according to the size of the GDP CV
              group_by(country, country_asGiven, continent) %>% #group data by country (both forms) and continent for the summarize function below - need to keep all three columns in summarize's output so that I can do further work with them all. This doesn't change the output of the summarize table, except to add columns to it.
              summarize(GDP_cv = coeff_var(gdpPercap)) %>% #summarize down to one column of the GDP coeff_var values to graph by
              ungroup() #ungroup prior to further analysis

```

Using the `arrange` function I can see that my reordering of the country column has been effective. In the first datatable output, using the country column in the default level order, Afghanistan comes first because it is the first country alphabetically. In the second datatable output, Senegal comes first because it has the lowest CV across its GDP data of all countries.
```{r factor_reord_arrange}

#this is the default order
gm_reord %>%
  arrange(country_asGiven) %>%
  DT::datatable()

#this is the reordered country column using CV(gdpPercap)
gm_reord %>%
  arrange(country) %>%
  DT::datatable()

```


To make the example graphs more readable, I first filtered it down to countries in Europe, but this would be applicable if you were keeping all countries, as well. Since I dropped countries by filtering down to Europe, this required me to drop unused factor levels:
```{r factor_reord_filter}

#filter and drop unused factor levels, in order to make graph more legible.
gm_reord_drop <- gm_reord %>%
                  filter(continent == "Europe") %>% #filter to Europe only
                  droplevels() #drop unused factor levels, to get rid of non-European countries and continents
                  
```


This first graph has its countries out of order relative to their summary data values, and it is hard to interpret:
```{r factor_reord_badGraph}

#this graph is hard to interpret:
gm_reord_drop %>%
  ggplot() +
    geom_col(aes(x = country_asGiven, #x axis is the non-reordered country levels, which are still alphabetical
                 y = GDP_cv)) + #bar size is based on max GDP for each country
    coord_flip() + #flip the graph to be displayed horizontally
    theme_bw() + #nicer black and white theme
    labs(x = "Countries, default alphabetical order", #explain why country axis seems out of order
         y = "Coefficient of variation in GDP per capita") #make better data label

```

This next graph is based on the same data, but uses the column of reordered countries, and is therefore more easily interpreted. Ireland has the highest variation in its GDP data, while Switzerland had the most stable data according to the coefficent of variation metric.
```{r factor_reord_goodGraph}

#this graph is easy to intepret:
gm_reord_drop %>%
  ggplot() +
    geom_col(aes(x = country, #x axis is the reordered country levels based on max(GDP)
                 y = GDP_cv)) + #bar size is based on max GDP for each country
    coord_flip() + #flip the graph to be displayed horizontally
    theme_bw() + #nicer black and white theme
    labs(x = "Countries, ordered by data values", #explain what country axis is showing
         y = "Coefficient of variation in GDP per capita") #make better data axis label

```


## Exercise 3: File input/output (I/O)

__Task:__ _Experiment with at least one of:_

    write_csv()/read_csv() (and/or TSV friends),
    saveRDS()/readRDS(),
    dput()/dget().

_You are expected to create something new, probably by filtering or grouped-summarization of your dataset (for e.g., Singer, Gapminder, or another dataset), export it to disk and then reload it back in using one of the packages above. You should use here::here() for reading in and writing out._

_With the imported data, play around with factor levels and use factors to order your data with one of your factors (i.e. non-alphabetically). For the I/O method(s) you chose, comment on whether or not your newly created file survived the round trip of writing to file then reading back in._


#### __My work__

I will do this exercise by looking at the relative variation in life expectancy, again summarized by the coefficient of variation (CV) metric. In this code chunk below, I am getting the CV for each country's life expectancy data over the 1952 to 2007 dataset period. I have grouped by both country and continent to keep two factors in play for the final step after writing and reading the object via RDS files.
```{r ex3_makeNew}

gm_summary <- gapminder %>%
                group_by(country, continent) %>%
                summarize(lifeExp_cv = coeff_var(lifeExp)) %>%
                ungroup()

```

In this code chunk below, I need to use the here::here package to write and read the file I want to create in a consistent way:
```{r ex3_here}

#this tells me that my root directory is my 'stat545-hw-trevor-baker' folder. if someone else opened this file from that folder, the path leading up to the 'stat545-hw-trevor-baker' folder would be different, but "here" would take care of writing system-compatible paths for that person as well:
here::here() 

#this tells me why 'here' has determined that to be my root directory -- it is because of the .Rproj file in that folder:
here::dr_here()

#this sets the full file path and name that I will write to, which I want to be in my Assignment_05 folder, two levels up from the 'here::here' root folder
file.loc <- here::here("Assignments","Assignment_05","hw05-gm_summary.RDS")

#this is the file location I will write to on my local machine:
print(file.loc)

```

The work above to set up the `file.loc` variable that holds my full file path and name can now be used to write and read the file in a reproducible way:
```{r ex3_io}

#save the file using saveRDS()
saveRDS(gm_summary, file.loc)

#just to prove that the file is really read in, I will read it to a new object name 'gm_input'
gm_input <- readRDS(file.loc)

#this new file is identical to the original, as proven by identical() returning TRUE:
identical(gm_input, gm_summary)

```
__I was successful in saving the file and then reading it back!__



Now I can manipulate the new tibble by filtering down to just Asia and adjusting the order of its factor levels. Here I'm aiming to organize the country levels according to the coefficient of variation, but I really want to have Cambodia first in my graph display.
```{r ex3_play}

#manipulate factor levels
gm_manip <- gm_input %>%
              filter(continent=="Asia") %>% #Asia only
              mutate(country = forcats::fct_reorder(country, lifeExp_cv, max)) %>% #order country levels by the CV values
              mutate(country = forcats::fct_relevel(country, "Cambodia")) #but put Cambodia first in the country levels

#check that this worked to have Cambodia first by checking the first six levels using 'head()':
gm_manip$country %>% levels() %>% head()

#make a nice plot showing the reordered factor levels
lifeExp_plot <- gm_manip %>%
                  ggplot(aes(x = country, 
                             y = lifeExp_cv, 
                             fill = country=="Cambodia")) + #fill Cambodia differently than other countries to further highlight it
                    geom_col() +
                    coord_flip() + #flip so that graph is displayed horizontally
                    theme_bw() + # nicer black and white theme
                    theme(legend.position = "none") + #remove the legend, not informative, just shows Cambodia = TRUE/FALSE
                    labs(y= "Coefficient of variation in life expectancy", #make clean x axis title
                         x = "") #remove label on country axis, redundant

#display the plot
lifeExp_plot

```


## Exercise 4: Visualization design

For this exercise, I am going to use a plot I made back in assignment 1. I looked at the relationship between GDP and life expectancy for the entire dataset, and see if there was a certain level of wealth at which the associated gains in life expectancy became muted. Without knowledge of `dplyr` and `ggplot2`, the process was more difficult than it could have been and the graph doesn't look as nice as what I can do now.

Below on the left is my old graph done wth base R plotting, and on the right is my new version of the graph made with `ggplot2`. 

- __Note:__ _this code is written in the order of (1. store new graph to object, 2. make old graph, 3. display new graph from object) because my old base R plot can't be saved to an object within an Rmd file. If I want the old one on the left and the new one on the right within a two-column display, the right panel will be filled entirely with my new graph's code if I try to code it directly after the old graph. As it is, the call to the new.gr object is still in that right panel, but it allows enough space for the graph to be displayed under it. If my old graph were a ggplot2 plot, this ugly code could be avoided by saving them both as objects and calling them in only two lines._
```{r ex4_new, warning=F}

new.gr <- gapminder %>%
            ggplot(aes(x = gdpPercap, 
                       y = lifeExp)) +
              geom_point( alpha=0.2, #make points transparent
                          size=2 ) + #make them a little larger
              geom_smooth( method = "lm", #linear model for the smoothed line
                           formula = y ~ log(x)) + #formula for the line- this actually makes a log relationship which is what I want, not linear
              scale_x_continuous(lim = c(0,50000), #limit the x-axis towards the low-GDP section in which the asymptote is approached
                                 labels = scales::dollar_format() ) + #put x axis labels in dollar format
              ylim(c(0,90)) +
              labs( x = "GDP per capita, ($ US, inflation-adjusted)", #make nice x axis label
                    y = "Life expectancy", #nice y axis label
                    title = "Life expectancy increases correlated to rising GDP seem to show diminishing returns",
                    caption = "Data source: R gapminder package, http://www.gapminder.org/data") + #tell the user what I'm trying to show
              theme_bw() + #use the nicer black and white theme
              theme(plot.title = element_text(size = 10, #change title size
                                              face = "bold"), #make title bold
                    plot.caption = element_text(size = 7), #make caption text smaller
                    axis.title = element_text( face = "bold") ) #make axis labels bold
```


<div class="twoC">
```{r ex4_old_and_display, warning=F}
#shorten name of gapminder dataset (lazy!)
gm <- gapminder

#need labels for legend
leg.labels <- levels(as.factor(gm$continent)) 

#create the plot with a legend
{ # these brackets are so the legend is written to the same image as the plot
  
  #create plot
  plot(gm$gdpPercap, #x variable
       gm$lifeExp,  #y variable
       xlim=c(0,50000), #x axis limits
       ylab="Life expectancy (years)",
       xlab="GDP per capita",
       main="Life expectancy vs GDP by continent",
       col=as.factor(gm$continent)) #colour points by continent
  
  #create legend in bottom right
  legend("bottomright", pch=21, pt.bg=1:length(leg.labels),
       legend=leg.labels ) 

} #end brackets to keep it together


new.gr #display the new graph

```
</div>
<div class="clearer"></div>


#### A comparison of the two graphs:
 - The quality of the image is higher in the new graph; it looks more modern and sharp.
 - In my original graph, I didn't really like how the hollow markers looked. but I couldn't use solid filled points because there was too much overlap In the new graph, the use of the _alpha_ argument allowed points to not overshadow eachother as much.
 - Because I didn't colour points by continent in the new graph, I didn't need a legend at all. If I did need a legend, then `ggplot2` would have done it for me automatically, saving the steps of manually creating it in my original code.
 
#### Things I did differently in the new graph:
 - I added a fitted logarithmic line to guide the viewer's eye to the trend I am trying to show.
 - I did not colour points by continents because the intention of the graph is to show the effect of GDP on life expectancy and extra colours distract the viewer's attention from what I want them to see.
 - I started the y-axis at zero, so as to give a better sense of proportion to the life expectancy increases. If I let the y limits be set automatically, starting at about 20, the increases in life expectancy appear more exaggerated.
 - I changed the figure title so that it better reflects the graph's purpose.
 - I added a caption for the data source, which I found in `?gapminder`.
 - I removed the redundant "(years)" part of the y-axis label.
 - I improved the y-axis label by clarifying the units of GDP.
 - The x-axis tick marks are labelled with dollar signs.
 
#### Things I thought about doing differently in the new graph but chose not to:
 - I considered putting a log x-axis with `scale_x_log10()`. It would declutter points and spread them out more, but I decided not to as the point I was originally trying to address was whether there is a certain GDP level that life expectancy gains are muted at. The linear x-axis shows the levelling of the trend better, and also allows easier reading of the GDP values.
 - I considered to turn off the x-axis limits, but the points past $50000 (all from Kuwait) are a distraction to the trend I am trying to highlight. The Kuwait points are interesting but they are a unique and separate story.


## Exercise 5: Writing figures to file

In order to save my plot in a reproducible way, I will use the `here` package to do so. I will save the last plot I made in exercise 4 by not specifying the `plot` argument within `ggsave`.
```{r ex5_save_last, warning = F, message = F}

#set file saving location with here::here
file.loc.ex5plot1 <- here::here("Assignments", "Assignment_05", "ex5_plot-GDP_x_lifeExp.png")

#because I am not specifying the plot argument, this will save the last plot I made
ggsave(filename = file.loc.ex5plot1, 
       device = "png")

```

Sometimes, though, I won't want to save the last plot I made and will need to use the `plot` argument of `ggsave`, as in the example below where I call in the plot object I made back in exercise 3. Just to mix it up, I will save this plot in pdf format (vector) instead of png format (raster), and I will save one copy in wide format and one in narrow format.
```{r ex5_save_earlier, warning = F, message = F}

#save the exercise 3 plot using here::here and ggsave(), make it wider
file.loc.ex5plot2_wide <- here::here("Assignments", "Assignment_05", "ex5_plot-lifeExp_CV_Asia-wide.pdf")
ggsave(filename = file.loc.ex5plot2_wide,
       plot = lifeExp_plot,
       device = "pdf",
       width = 14,
       height = 6)

#save the exercise 3 plot but make it tall and narrow
file.loc.ex5plot2_tall <- here::here("Assignments", "Assignment_05", "ex5_plot-lifeExp_CV_Asia-tall.pdf")
ggsave(filename = file.loc.ex5plot2_tall,
       plot = lifeExp_plot,
       device = "pdf",
       height = 12,
       width = 6)

```


Now I can embed the files in my markdown report via links. Here is [the wide one](ex5_plot-lifeExp_CV_Asia-wide.pdf) and here is [the tall and narrow one](ex5_plot-lifeExp_CV_Asia-tall.pdf).
